Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-05-06 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: implemented Anuradha Gupta`s comments

Several null tests have been proposed to probe the nature of gravity with GW signals~\cite{TheLIGOScientific:2016src,Yunes:2016jcc,Abbott:2018lct,LIGOScientific:2019fpa,Abbott:2020jks,LIGOScientific:2021sio}. They include tests of GW generation~\cite{Arun:2006yw,Yunes:2009ke,Li:2011cg,Agathos:2013upa,Mehta:2022pcn}, where deviations in the post-Newtonian (PN) coefficients in the inspiral, and phenomenological coefficients in the plunge and merger stages can be bounded; tests of GW propagation~\cite{Will:1997bb}, which allow us to set upper limits on coefficients entering generalized dispersion relations, including the Compton wavelength associated to the
mass of the graviton; tests of the polarization of gravitational radiation~\cite{Will:2014kxa}, for which more than
two GW detectors are needed to set statistically significant bounds, and tests of the remnant properties~\cite{Meidam:2014jpa,Carullo:2018sfu,Brito:2018rfr,Carullo:2019flw,Isi:2019aib,Ghosh:2021mrv,Carullo:2021dui}
during in the post-merger. So far, none of these null tests have reported any deviation from GR.

@@

\begin{table}[t]
\begin{threeparttable}
\begin{tabular}{c | c c}
\hline \hline
Theory & Constraint & This work \\
\hline
EdGB        & $\ell_{\rm EdGB} \leqslant 1.18$~km~(GW)~\cite{Lyu:2022gdr} & -- \\
dCS         & $\ell_{\rm dCS} \leqslant 8.5$~km~(EM+GW)~\cite{Silva:2020acr}  & $\ell_{\rm dCS} \leqslant 38.7$~km \\
cubic EFT   & -- & $\ell_{\rm cEFT} \leqslant 38.2$~km \\
quartic EFT & $\ell_{\rm qEFT} \leqslant 150$~km~\cite{Sennett:2019bpc}  & $\ell_{\rm qEFT} \leqslant 51.3$~km~\tnote{a}
\\
\hline \hline
\end{tabular}
\caption{Summary of the upper limits, at $90\%$ credible level,
on the length-scale of the modified gravity theories under investigation.
They were obtained combining Bayesian-inference results from GW150914 and GW200129.
Current constraints on $\ell_{\rm th}$ are also listed.}
\begin{tablenotes}
\item[a]{\footnotesize We notice that our result for the quartic EFT of GR is only in marginal tension with our hard cut-off scale for the
validity of the theory, and hence we do still quote it here (see Secs.~\ref{sec:remarks} and~\ref{sec:results_qeft}
for details)}
\end{tablenotes}
\label{tab:bound_summary}
\end{threeparttable}
\end{table}

@@

For $N$ independent GW observations $\{d_j\}$, $j=1,...,N$, each characterized
by a PDF $P_j(\ell_{\rm th} \vert d_j,\mathcal{H})$, the
joint posterior can be written as:
%
\begin{equation}
P(\ell_{\rm th} | \{d_j\},\mathcal{H}) = p(\ell_{\rm th}) \prod_{j=1}^{N} \frac{P_j(\ell_{\rm th} | d_j,\mathcal{H})}{p_j(\ell_{\rm th}|\mathcal{H})}\,.
\label{eq:cumulative_dist_ell}
\end{equation}
%
where $p_j(\ell_{\rm th} | \mathcal{H})$ are the priors used for each observation, and $p(\ell_{\rm th} | \mathcal{H})$ is an overall prior.
%
Since we assume a uniform prior on $\ell_{\rm th}$, the joint posterior is
equal to the joint likelihood. Hereafter, we will drop the explicit usage of
$\mathcal{H}$.
%
%In some cases, we found it necessary to work with different prior ranges across
%different events, within the context of a given theory.
%
%In this situation, when combining the posteriors we use chose as our overall
%prior the widest one among the individual events.

@@

In Sec.~\ref{sec:results}, we enumerate through the different theories and
outline the main results. Whenever possible, we also combine results from
multiple events to obtain the strongest possible bound on $\ell_{\rm th}$.

@@

%
As an additional check, to verify the robustness of our constraint, we show in
Fig.~\ref{fig:corner_plot}, the final spin $\chi_{\rm f}$ and remnant mass
$M_{\rm f}$ for GW150914 for GR and dCS gravity.
%
We see that our \pSEOB~waveform model does not introduce substantial changes
to the GR estimates on these parameters, as required by the ParSpec expansion
(see discussion in Sec.~\ref{sec:review_parspec}).
%
In fact, we observed no bias on the estimation of $\Mf$ and $\chi_{\rm f}$
for all theories considered here.
%
For completness, in Appendix~\ref{app:big_contour} we also
show how all other intrinsic parameter parameters remain unbiased.

>>>>

Several null tests have been proposed to probe the nature of gravity with GW signals~\cite{TheLIGOScientific:2016src,Yunes:2016jcc,Abbott:2018lct,LIGOScientific:2019fpa,Abbott:2020jks,LIGOScientific:2021sio}. They include tests of GW generation~\cite{Arun:2006yw,Yunes:2009ke,Li:2011cg,Agathos:2013upa,Mehta:2022pcn}, where deviations in the post-Newtonian (PN) coefficients in the inspiral, and phenomenological coefficients in the plunge and merger stages can be bounded; tests of GW propagation~\cite{Will:1997bb}, which allow us to set upper limits on coefficients entering generalized dispersion relations, including the Compton wavelength associated to the
mass of the graviton; tests of the polarization of gravitational radiation~\cite{Will:2014kxa}, for which more than
two GW detectors are needed to set statistically significant bounds, and tests of the remnant properties~\cite{Meidam:2014jpa,Carullo:2018sfu,Brito:2018rfr,Carullo:2019flw,Isi:2019aib,Ghosh:2021mrv,Carullo:2021dui}
\redsout{during} \hs{in the post-merger}. So far, none of these null tests have reported any deviation from GR.

@@

\begin{table}[t]
\begin{threeparttable}
\begin{tabular}{c | c c}
\hline \hline
Theory & Constraint & This work \\
\hline
EdGB        & $\ell_{\rm EdGB} \leqslant 1.18$~km~(GW)~\cite{Lyu:2022gdr} & -- \\
dCS         & $\ell_{\rm dCS} \leqslant 8.5$~km~(EM+GW)~\cite{Silva:2020acr}  & $\ell_{\rm dCS} \leqslant 38.7$~km \\
cubic EFT   & -- & $\ell_{\rm cEFT} \leqslant 38.2$~km \\
quartic EFT & $\ell_{\rm qEFT} \leqslant 150$~km~\cite{Sennett:2019bpc}  & $\ell_{\rm qEFT} \leqslant 51.3$~km~\tnote{a}
\\
\hline \hline
\end{tabular}
\caption{Summary of the upper limits, at $90\%$ credible level,
on the length-scale of the modified gravity theories under investigation.
They were obtained combining Bayesian-inference results from GW150914 and GW200129.
Current constraints on $\ell_{\rm th}$ are also listed.}
\begin{tablenotes}
\item[a]{\footnotesize We notice that our result for the quartic EFT of GR is only in marginal tension with our hard cut-off scale for the
validity of the theory, and hence we do still quote it here (see Secs.~\ref{sec:remarks} and~\ref{sec:results_qeft}
for details)\hs{.}}
\end{tablenotes}
\label{tab:bound_summary}
\end{threeparttable}
\end{table}

@@

For $N$ independent GW observations $\{d_j\}$, $j=1,...,N$, each characterized
by a PDF $P_j(\ell_{\rm th} \vert d_j,\mathcal{H})$, the
joint posterior can be written as:
%
\begin{equation}
P(\ell_{\rm th} | \{d_j\},\mathcal{H}) = p(\ell_{\rm th}) \prod_{j=1}^{N} \frac{P_j(\ell_{\rm th} | d_j,\mathcal{H})}{p_j(\ell_{\rm th}|\mathcal{H})}\,.
\label{eq:cumulative_dist_ell}
\end{equation}
%
where $p_j(\ell_{\rm th} | \mathcal{H})$ are the priors used for each observation, \redsout{and} $p(\ell_{\rm th} | \mathcal{H})$ is an overall prior\hs{, and we assume that the value of $\ell_{\rm th}$ is shared among all events.}
%
Since we assume a uniform prior on $\ell_{\rm th}$, the joint posterior is
equal to the joint likelihood. Hereafter, we will drop the explicit usage of
$\mathcal{H}$.
%
%In some cases, we found it necessary to work with different prior ranges across
%different events, within the context of a given theory.
%
%In this situation, when combining the posteriors we use chose as our overall
%prior the widest one among the individual events.

@@

In Sec.~\ref{sec:results}, we enumerate through the different theories and
outline the main results. Whenever possible, we also combine results from
\redsout{multiple} \hs{both} events to obtain the strongest possible bound on $\ell_{\rm th}$.

@@

%
As an additional check, to verify the robustness of our constraint, we show in
Fig.~\ref{fig:corner_plot}, the final spin $\chi_{\rm f}$ and remnant mass
$M_{\rm f}$ for GW150914 for GR and dCS gravity.
%
We see that our \pSEOB~waveform model does not introduce substantial changes
to the GR estimates on these parameters, as required by the ParSpec expansion
(see discussion in Sec.~\ref{sec:review_parspec}).
%
In fact, we observed no bias on the estimation of $\Mf$ and $\chi_{\rm f}$
for all theories considered here.
%
For completness, in Appendix~\ref{app:big_contour} we also
show how all other intrinsic \redsout{parameter} parameters remain unbiased.
